---
title: "Technical Execution Roadmap — Phase 1: Data Analysis"
subtitle: "Sentiment Analysis Project | JEMIB IT&BI Area · Spring 2026"
author: "Colombini Francesco & Hosen Jabed"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
    number_sections: true
---

```{r global-setup, include=FALSE}
#Environment & Dependencies
knitr::opts_chunk$set(
  echo        = TRUE,
  warning     = FALSE,
  message     = FALSE,
  fig.align   = "center",
  fig.width   = 10,
  fig.height  = 5.5,
  cache       = FALSE
)

rm(list=ls())
# Use project root for relative paths and allow a private local data folder.
knitr::opts_knit$set(root.dir = normalizePath(".", 
						winslash = "/", mustWork = TRUE))
private_data_dir <- Sys.getenv("PRIVATE_DATA_DIR", unset = "data")
```

------------------------------------------------------------------------

```{r libraries}

# Declare and load all required libraries.
library(tidyverse)    # dplyr, ggplot2, stringr, forcats, purrr
library(lubridate)    # robust datetime construction
library(tidytext)     # unnest_tokens(), stop_words, bind_tf_idf()
library(knitr)        # kable()
library(kableExtra)   # styled tables
library(textclean)
library(SnowballC)

# ── Statistical testing
library(rstatix)      # chi_sq_test(), cramer_v(), effect size helpers

```

------------------------------------------------------------------------

# Step 1 — Data Ingestion & Schema Validation

Both CSVs were loaded, enforcing correct column types, and verifying the schema matches the project specification.

```{r data-ingestion}
library(tidyverse)
SA_test <- as_tibble(read_csv2("sentiment_analysis_test.csv"))
SA_train <- as_tibble(read_csv2("sentiment_analysis_train.csv"))
summary(SA_train)
head(SA_train)
```

```{r schema-table, echo=FALSE}
tibble(
  Column         = c("Year","Month","Day","date","time_of_tweet","text","sentiment","platform"),
  Type           = c("int","int","int","Date","factor(3)","chr","factor(3)","factor(3)"),
  Expected_Range = c("2010–2023","1–12","1–31","derived","morning/noon/night","free text","neg/neu/pos","FB/IG/TW"),
  Phase_1_Role   = c("temporal trend","seasonality","granularity","timeline axis","posting rhythm","NLP corpus","target variable","stratification variable")
) |>
  kable(caption = "Schema Reference Table — Training Set") |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

```{r}
slang_map <- c(
  "u" = "you",
  "ur" = "your",
  "imo" = "in my opinion",
  "idk" = "i do not know",
  "omg" = "oh my god",
  "btw" = "by the way"
)

clean_text <- function(x){
  x %>%
    str_to_lower() %>%
    str_replace_all("http[s]?://\\S+|www\\.\\S+", " URL ") %>%
    str_replace_all("@\\w+", " USER ") %>%
    str_replace_all("#(\\w+)", " \\1 ") %>%             # keep hashtag word
    str_replace_all("([!?.,])\\1{1,}", "\\1\\1") %>%   # cap repetition
    str_replace_all("\\b(u|ur|imo|idk|omg|btw)\\b", ~ slang_map[.x]) %>%
    str_squish()
}

cleaning_table <- function(data){
  data$Year <- as.integer(as.character(data$Year))
  data$Month <- as.integer(as.character(data$Month))
  data$Day <- as.integer(as.character(data$Day))
  data$`Time of Tweet` <- factor(data$`Time of Tweet`,  
                                 levels = c("morning","noon","night"))
  data$sentiment <- factor(data$sentiment,
                           levels = c("negative", "neutral", "positive"))
  data$text <- clean_text(data$text)
  data$Platform <- factor(data$Platform)
  data
}

SA_train_cleaned <- cleaning_table(SA_train)
SA_train_cleaned <- SA_train_cleaned %>%
  mutate(date = make_date(Year, Month, Day),
    .after = Day)

cat("Train set")
summary(SA_train_cleaned)

SA_test_cleaned <- cleaning_table(SA_test)
SA_test_cleaned <- SA_test_cleaned %>%
  mutate(date = make_date(Year, Month, Day),
    .after = Day)

cat("Test set")
summary(SA_test_cleaned)
```

------------------------------------------------------------------------

# Step 2 — Structural Data Quality Audit

A data quality procedure must be ensured to justify every decision made in the following steps.

```{r data-quality}
missing_map <- SA_train_cleaned %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), 
               names_to = "column", 
               values_to = "n_missing") %>%
  mutate(pct_missing = round((n_missing / nrow(SA_train)/100),3))

missing_map
```

```{r duplicates}
n_exact_dupes <- sum(duplicated(SA_train_cleaned$text))
cat("Exact text duplicates detected:", n_exact_dupes, "\n")

#remove duplicates, probably from different platforms
SA_train_cleaned <- SA_train_cleaned %>%
  distinct(text, .keep_all = TRUE)
cat("Removed ", n_exact_dupes, "duplicates\n")

# Flag suspiciously short posts (< 3 words — likely noise or retweet stubs)
SA_train_cleaned <- SA_train_cleaned |>
  mutate(
    word_count  = str_count(text, "\\S+"),
    flag_short  = word_count < 3
  )

SA_train_cleaned %>%
  count(flag_short) %>%
  kable(caption = "Words with < 3 words") %>%
  kable_styling(bootstrap_options = "condensed", full_width = FALSE)
```

```{r date-validation}
# Temporal range validation
SA_train_cleaned |>
  summarise(
    earliest_post = min(date, na.rm = TRUE),
    latest_post   = max(date, na.rm = TRUE),
    n_years       = n_distinct(Year),
    n_invalid_dates = sum(is.na(date))
  ) %>%
  kable(caption = "Temporal Integrity Check") |>
  kable_styling(bootstrap_options = "condensed", full_width = FALSE)
```

------------------------------------------------------------------------

# Step 3 — Data Cleaning & Text Preprocessing Pipeline

Social media language is adversarial to standard NLP tools, URLs are noise, emojis carry semantic weight and contractions fragment token counts.

```{r text-cleaning-pipeline}
SA_train_cleaned <- SA_train_cleaned %>%
  mutate(
    text_raw = text,
    text = str_replace_all(text, "&amp;", "&"),
    text = str_replace_all(text, "&lt;", "<"),
    text = str_replace_all(text, "&gt;", ">"),
    text = str_remove_all(text, "https?://\\S+|www\\.\\S+"),
    text = str_remove_all(text, "@\\w+"),
    text = str_replace_all(text, "#(\\w+)", "\\1"),
    text = textclean::replace_contraction(text), 
    text = textclean::replace_emoji(text),
    text = textclean::replace_internet_slang(text),
    text = str_remove_all(text, "[^[:alpha:][:space:]]"),
    text = str_squish(text),
    text = str_to_lower(text)
  )


# Quick sanity check: display before/after for 3 rows
SA_train_cleaned %>%
  slice(1:3) %>%
  select(text_raw, text) %>%
  kable(caption = "Cleaning Pipeline — Before vs After (sample)") %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = FALSE)
```

```{r stopwords}
# remove the words with zero semantic meaning on their own
custom_stopwords <- bind_rows(
  get_stopwords(source = "snowball"),
  tibble(word = c("rt","amp","via","http","https","just","get","got",
                  "like","im","dont","cant","thats","youre","ive",
                  "ill","its","youll"),
         lexicon = "custom_jemib")
)

# Tokenise, remove stopwords, stem
tokens <- SA_train_cleaned %>% #first drop the useless column
  select(Year, date, `Time of Tweet`, sentiment, Platform, text) %>%
  unnest_tokens(word, text) %>% #explodes the rows
  anti_join(custom_stopwords, by = "word") %>% #drops the stop words
  filter(str_length(word) > 2) %>% # drop 1–2 char residuals
  mutate(word_stem = wordStem(word))  # Stemming on the remaining words

cat("Total tokens after cleaning:", nrow(tokens), "\n")
cat("Unique stems:", n_distinct(tokens$word_stem), "\n")
```

------------------------------------------------------------------------

# Step 4 — Univariate EDA: Target Variable (Sentiment Distribution)

Class imbalance is not only a modelling concern, because it can distort the story told in the dashboard. If 40% of posts are neutral, a chart that shows equal bars is misleading. Quantifying imbalance here provides the statistical foundation for any sampling strategy.

```{r sentiment-dist}
# Frequency table with proportions
sentiment_freq <- SA_train_cleaned %>%
  count(sentiment) %>%
  mutate(percent = paste0(round(n / sum(n) * 100, 1), "%"))

sentiment_freq %>%
  kable(caption = "Sentiment Class Distribution — Training Set (n = 326)") %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = FALSE)
```
